---
title: "AUGglmmTMB"
output: rmarkdown::github_document
---

## Overview

`AUGglmmTMB` is an R package for the analysis of binomial mixed models in settings with sparse and clustered data. It provides methods for stable estimation in the presence of separation and boundary estimates of random effects covariance matrices.

The package implements penalized likelihood approaches based on extensions of Firth-type penalty for the fixed effects and Inverse Wishart type penalty for the random effects covariance matrix through data augmentation techniques as presented in Košuta et al. These methods ensure finite and interpretable parameter estimates while preserving desirable statistical properties such as parameterization invariance.

## Motivation

In many applied settings, particularly in clinical and epidemiological research, data are sparse and structured hierarchically. Standard maximum likelihood estimation in binomial mixed models may fail due to:

- separation in binary or count outcomes,
- rare events or rare exposures,
- boundary estimates of variance components.

These issues lead to unstable estimation and invalid inference. `AUGglmmTMB` provides practical tools to address these challenges within a unified framework.

## Features

- Penalized likelihood estimation for GLMMs based on Firth-type corrections  
- Data augmentation approaches for stable estimation  
- Handling of separation in clustered data  
- Regularization of random effects covariance parameters  
- Compatibility with `glmmTMB` workflows  

## Key Functions

- `AUGglmmTMB()`: fits penalized binomial mixed models using iterative algorithms based on Košuta et al. The function supports penalties on both fixed and random effects and provides stable estimation in the presence of separation and boundary estimates.

- `mpl_fitter()`: fits penalized binomial mixed models using iterative algorithms based on Košuta et al. The function supports penalties on both fixed and random effects and provides stable estimation in the presence of separation and boundary estimates.

- `get_psi()`: estimates penalty parameters for the random-effects covariance structure using an iterative procedure consistent with the penalized likelihood framework.

- `get_data_plot_cloglik()`: evaluates the conditional log-likelihood over a grid of shrinkage parameters for the random-effects covariance matrix, supporting data-driven selection of the penalty strength obtained using `get_psi()`.

## Installation

You can install the development version from GitHub:

`install.packages("remotes");remotes::install_github("rokblagus/AUGglmmTMB")`

## Examples

The functionality of the package is illustrated using a dataset on European passerine birds, originally published by Bandelj et al. The data used in their study included additional sampling to address the separation issue. In contrast, the dataset `data(birds)` provided with this R package corresponds to the original dataset, prior to this additional sampling.

First, we fit the model using ML, as implemented in the R package `glmmTMB`, using the default settings.

```{r}
library(glmmTMB)
library(AUGglmmTMB)
data(birds)
fit_ml<-glmmTMB(parasites~migration+food+(migration|phylogenetic)+(1|species),data=birds,family=binomial(link = "logit"))
summary(fit_ml)

```

Due to the separation issue, the coefficient for `foods` is unreasonable. The estimated random effects covariance matrix at the level of `phylogenetic` is on the boundary of the parameter space, as evident by the correlation parameter `Corr` estimated at -1. Furtheremore, the standar errors cannot be computed for any of the model's parameters.

```{r}
fit_ml$sdr
```


To solve these issues, we fit the model using MPL, using the default penalty for the fixed effects penalty, inverse-Wishart penalty with $\nu=3$ and $\Psi=I_2$ ($2\times 2$ identity matrix) for the random effects covariance matrix at the level of `phylogenetic` and no penalty for the random effects covariance by `species`. We use Algorithm 1 from Košuta et al.

```{r}
fit_mpl_1<-AUGglmmTMB(parasites~migration+food+(migration|phylogenetic)+(1|species),data=birds,link = "logit",
                penOpt = AUGglmmTMBPenalty(autrepen =FALSE,nu=list(3),psi=list(diag(1,2,2))),
                      control=AUGglmmTMBControl(fit_pGLM = FALSE,maxiter = 50,tol=1e-5,save_coef=TRUE)   )
summary(fit_mpl_1$fit$fit)

```


The coefficient for `foods` is now more reasonable, with the random effects covariance matrix at the level of `phylogenetic` estimated away from the boundary. Unlike for ML, the standard errors for model's parameters can now be computed. However, there is still an issue with the SE for one of the random effects parameters, implying that the chosen prior for the random effects is not optimal.

```{r}
fit_mpl_1$fit$fit$sdr
```

Therefore, we use the data-driven procedure proposed by Košuta et al., to determine the penalty parameters for the random effects penalty. 

```{r}
fit_mpl_2<-AUGglmmTMB(parasites~migration+food+(migration|phylogenetic)+(1|species),data=birds,link = "logit",
                penOpt = AUGglmmTMBPenalty(autrepen =TRUE,nu=list(3),plot=TRUE,ntaus=20),
                      control=AUGglmmTMBControl(fit_pGLM = FALSE,maxiter = 50,tol=1e-5,save_coef=TRUE)   )
```


The plot produced by `plot=TRUE,ntaus=20`, shows the conditional likelihood as a function of $\tau$, with the blue vertical line denoting the chosen value of $\tau$. Using this data-driven penalty 
```{r}
fit_mpl_2$optre
```
the standard errors are now available for all model's parameters
```{r}
fit_mpl_2$fit$fit$sdr
summary(fit_mpl_2$fit$fit)
```

We now use a computationally more efficient approximation of MPL, based on a single iteration of Algorithm 2 in Košuta et al.
```{r}
fit_mpl_3<-AUGglmmTMB(parasites~migration+food+(migration|phylogenetic)+(1|species),data=birds,link = "logit",
                penOpt = AUGglmmTMBPenalty(autrepen =TRUE,nu=list(3),plot=TRUE,ntaus=20),
                      control=AUGglmmTMBControl(fit_pGLM = TRUE,maxiter = 1,tol=1e-5,save_coef=TRUE)   )
```

which gives very similar results than the exact implementation based on Algorithm 1: the difference between the two procedures relative to the SE obtained by the exact procedure is smaller than 1% of the estimated SE for all model's parameters
```{r}
(fit_mpl_2$fit$fit$sdr$par.fixed-fit_mpl_3$fit$fit$sdr$par.fixed)/sqrt(diag(fit_mpl_2$fit$fit$sdr$cov.fixed))*100
fit_mpl_3$fit$fit$sdr
summary(fit_mpl_3$fit$fit)
```


To implement a penalty on both random effects, we can e.g. use
```{r}
fit_mpl_4<-AUGglmmTMB(parasites~migration+food+(migration|phylogenetic)+(1|species),data=birds,link = "logit",
                penOpt = AUGglmmTMBPenalty(autrepen =FALSE,nu=list(3,1),psi=list(fit_mpl_3$optre$psi,matrix(1,1,1))),
                      control=AUGglmmTMBControl(fit_pGLM = TRUE,maxiter = 1,tol=1e-5,save_coef=TRUE)   )
```

where we used the data-driven penalty for the covariance matrix at the level of `phylogenetic` and the inverse-Wishart prior with $\nu=1$ and $\Psi=1$ for the covariance at the level of `species`.
```{r}
summary(fit_mpl_4$fit$fit)
fit_mpl_4$fit$fit$sdr

```



